---
title: Further testing of quickRareCurve
author: ~
date: '2019-01-01'
slug: further-testing-of-quickrarecurve
categories: []
tags: []
image:
  caption: ''
  focal_point: ''
---

After my post yesterday, documenting a faster parallelised version of the `rarecurve` function (`quickRareCurve`), I realised it'd be good to show a real world example using it on a reasonably large OTU table, to prove that it is indeed quicker than the original function. So, here we go.

```{r, warning = F, message = F, echo = F}
library(data.table)
library(vegan)
bac <- fread("bac16S.csv")
bac <- bac[Chimer_ref == "non_chimera_ref", ]

bac[, c("Chimer_ref", "Chimera_denovo") := NULL]

setDF(bac)
otuTable <- dcast(melt(bac, id.vars = "OTUId"), variable ~ OTUId)
setDT(otuTable)
```
```{r, warning = F, message = F}
# Starting with an OTU table in which rows are samples,
# Cols are OTUs/species
dim(otuTable)

# the first column of my data is the sample names
# so remember [, -1] to not include it
# Lets inspect sample sizes with a simple histogram
hist(rowSums(otuTable[, -1]), xlab = "Sample sizes")
```

In this example, our OTU table contains 48 samples and 5381 OTUs, plus a column with the sample names in.

```{r, warning = F, message = F, echo = F}
# install from ecolFudge package
library(devtools)
install_github("dave-clark/ecolFudge")

library(ecolFudge)
```

Now we can use `microbenchmark` again to compare the performance of the original `rarecurve` function to our faster parallel version, `quickRareCurve`. We will then plot the results using `ggplot2`. Warning: the code below will take some time to run!

```{r, warning = F, message = F, eval = F}
library(microbenchmark)

# benchmark the two functions with 3 replicates
testResult <- microbenchmark(
  rarecurve(otuTable[, -1]),
  quickRareCurve(otuTable[, -1]),
  times = 3)
```

```{r, echo = F}
testResult <- readRDS("quickRareCurveBenchmark.RDS")

```{r, warning = F, message = F}
library(ggplot2)

# convert from nanoseconds to minutes by dividing
# time by 6e+10
ggplot(testResult, aes(x = expr, y = time/6e+10)) +
  geom_boxplot() +
  labs(x = "Function", y = "Time (minutes)") +
  theme(axis.text = element_text(size = 16),
    axis.title = element_text(size = 18))
```

There you go, on a typical OTU table, the `quickRareCurve` function is far quicker, reducing the processing time from ~25 minutes to < 5 minutes. The more samples in your OTU table, and the more CPU cores available, the greater the increase in performance.
